---
csl: apa-cv.csl
bibliography: references.bib
suppress-bibliography: true
link-citations: false
citations-hover: false
---

# Workshop overview

```{r}
#| eval: false
#| echo: false
# https://www.andrewheiss.com/blog/2023/01/09/syllabus-csl-pandoc/#using-other-styles
# run manually
rbbt::bbt_update_bib(here::here("index.qmd"))
```

This web-book contains slides for the workshop 'Open Science Practices for Linguistic Research: Reproducible Analyses in R' given given by Daniela Palleschi at the [Summer School of Linguistics](https://ssol.ff.cuni.cz/) in Budweis, Czechia in August 2024. The tools discussed are specific to the R enviornment, but the concepts are universal and programming language agnostic. The materials are re-structured from a semester-length course on the same topic given at the Humboldt-Universität zu Berlin in the summer semester 2024, the materials for which are more exhaustive ([click here](https://daniela-palleschi.github.io/r4repro_SoSe2024/) to see course materials).

Workshop abstract:

> The Open Science movement began as an answer to the replication crisis and aims to encourage transparency across all stages of research. In this workshop, we will focus on practicing transparency in our analyses through reproducibility: what does it mean, why should we practice it, and how can we do it? We will focus on establishing and maintaining a reproducible, project-oriented workflow in the R environment. After the workshop, participants will be able to put reproducibility concepts and tools into practice, such as data management and documentation, R projects, and R packages developed specifically for reproducibility. The workshop assumes participants have at least basic familiarity with R and RStudio.

## Schedule

```{r}
#| echo: false
dplyr::tribble(
  ~"Session", ~"Topic(s)",
  "Day 1", "The state of reproducibility in linguistic research",
  "Day 2", "(i) Setting up a reproducible project-oriented workflow <br> (ii) Implementing literate and modular analyses",
  "Day 3", "Putting it into practice",
  "Day 4", "(i) Publishing analyses<br> (ii) Conducting peer code review"
) |> 
  knitr::kable()
```

## Workshop goals

We will discuss and implement the following:

- reproducibility rates in linguistics
- project-oriented workflow in R
  + with RProjects
  + folder structure
  + naming conventions
  + using project-relative filepaths with the `here` package
- literate programming
  + writing linear code
  + modular analyses
  + dynamic reports with Quarto
- sharing and checking our code
  + uploading code to an OSF repository
  + conducting a code review
  
What we will *NOT* cover:

- version control (e.g., git/GitHub)
- learning R/the RStudio environment
- how to appropriately analyse data (e.g., which analyses to use, etc.)
- how to produce tables and figures
- how to write a manuscript in Rmarkdown

What we *might* cover if there's interest and time:

- project-relative package management with the `renv` package

# Preparation

It is assumed you have at least some basic familiarity with R and R Studio. Please at the very least make sure you have the required software before Day 1.

## Software (before Day 1)

1. install or update R to version 4.4.0 (Puppy Cup, 2024-04-24)
    + *Disclaimer*: updating R can interfere with on-going R projects you are currently working on, most notably because you will need to re-install packages (and thus you may be installing more recent package versions which may break existing code). If you are currently in the middle of analysing some data, you may not want to update R right now. In this case, just make note of which version you're currently running (e.g., by running `R.version` in the Console)
    + having an R version from 2022.07 or later should suffice
2. install or update RStudio
    + I am currently using RStudio version 2023.12.1+402, as I encountered issues when updating to 2024.04.2+764 in April when it was released. As a rule of thumb, I update R and/or RStudio a few months after their initial release, and when I know I have time to fix any bugs that might pop up (i.e., I don't have a looming deadline)


## Additional steps (before Day 4)

Create an OSF account [here](https://osf.io/register) if you don't have one already

## How to navigate this website

Each topic is listed in the sidebar in chronological order. Three output formats are available, all with the same content:

1. HTML page
2. PDF of content (sub-optimally formatted)
3. RevealJS slides

The contents were formatted for the slide output. Tables and figures may be too large/small in HTML and PDF format (especially the latter). Each page of the website presents the HTML format. The other 2 formats can be viewed by clicking on their symbol under 'Other Formats' (right sidebar).

# Suggested readings (before, during, or after the workshop)

There is currently a wealth of literature on the topic of reproducibility, both in terms of meta-science reviews of rates of reproducibility and in terms of best-practice advice. Some reading I would suggest for a soft introduction into the latter would be:

- @nagler_coding_1995
- @bowers_how_2016
- @wilson_good_2017

For a book-length treatment of R-specific reproducible workflows:

- @rodrigues_building_nodate

And for an overview of R-specific data analysis suggestions, I recommend the following on-line resources:

- @bryan_what_nodate
- @bryan_chapter_nodate

For more general discussions on Open Science Practices:

- @kathawalla_easing_2021