---
title: "Building a reproducible workflow in R"
subtitle: "Project-oriented workflow"
author: "Daniela Palleschi"
institute: "Humboldt-Universität zu Berlin"
lang: en
date: 2024-08-21
date-format: "ddd MMM D, YYYY"
date-modified: last-modified
language: 
  title-block-published: "Workshop Day 1"
  title-block-modified: "Last Modified"
format: 
  html:
    number-sections: true
    number-depth: 2
    toc: true
    code-overflow: wrap
    code-tools: true
    embed-resources: false
  pdf:
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    footer: "SSOL 2024"
    include-in-header: ../../mathjax.html # for multiple
    output-file: R-workflow_slides.html
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    embed-resources: false
    fig-align: center
    fig-dpi: 300
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: ../../../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

```{r}
#| echo: false
source(here::here("functions", "print_image.R"))
```


```{r}
#| echo: false
#| eval: false
# run manually
rbbt::bbt_update_bib(here::here("slides", "day1", "R-workflow", "R-workflow.qmd"))
```

# Learning Objectives {.unnumbered .unlisted}

Today we will learn...

-   how to implement a project-oriented workflow in R
-   how to organise folders and file names
-   how to set up our first R project

# Building a reproducible workflow in R

- we now know some important principles of a reproducible workflow
  + and that 'reproducibility' is not black-and-white
  + but even the reproducibility spectrum is an oversimplification [@peng_reproducible_2011]

- some additional resources that provide a list of tips include:
  - @bowers_how_2016; @nagler_coding_1995; @wilson_good_2017; @corker_open_2022

## Broadening the reproducibilty spectrum

-   there are different levels of reproducibility
    -   the *bare minimum* is sharing the code and data
    -   *and* including session information:
        -   which operating system was used
        -   which software/package versions were used
-   going bigger:
    -   project-oriented workflow
    -   project-specific filepaths
    -   contained in a single project folder
-   we will be using RProjects to achieve this

## Project management

-   folder structure
-   project-relative file paths
-   appropriate documentation
    -   e.g., README
-   it's great to map out your project structure early on
    -   but it will grow as you go along
    -   reproducible principles facilitate adapting as it grows

## Literate programming

> Instead of imagining that our main task is to instruct a *computer* what to do, let us concentrate rather on explaining to *human beings* what we want a computer to do.

--- @Knuth_literate_1984, p. 97

::: columns
::: {.column width="100%"}
-   originally used to refer to writing programs
-   but also applies to analysis code
    -   especially if we're aiming for reproducibility
:::

::: {.column width="50%"}
-   main concepts:

    -   code is linear [this pre-dates @Knuth_literate_1984]
    -   informative but concise commenting
:::

::: {.column width="50%"}
-   main benefits:

    -   facilitates maintenance
    -   helpful for future-you, collaborators, etc.
:::
:::

## Documentation

-   metadata
    -   project README
    -   codebook/data dictionary

-   README should contain
    -   a project description
    -   relevant links
    -   description of folder structure

-   can be updated as the project develops

-   README.md files in GitHub/Lab are automatically used as a project description
    -   `.md` is a plain text document
    -   uses markdown syntax

## Version control (not covered in this workshop)

-   git: local tracking
-   useful for the analysis and writing phases
    -   but can be tricky for collaboration
-   GitHub/GitLab: remote tracking
    -   store your changes to your local git repository
    -   then push them to your remote repository
-   safe guards against local hardware/software issues
    -   lost or damaged computer or local files
-   and allows for collaboration or sharing

## Persistant (public) storage

-   GitHub/Lab are sub-optimal
    -   developer-focused
    -   typically lack thorough documentation/metadata
    -   not very user-friendly for non-users
-   OSF, Zenodo
    -   Open Science-focused
    -   can be linked to a GitHub/Lab repository
    -   facilitate thorough documentation
    -   user-friendly

## Writing (not covered in this workshop)

-   dynamic reports with Markdown syntax

    -   e.g., Rmarkdown, Quarto
    -   integration of data, code, and prose
        -   facilitates cross-referencing within document
        -   integration of citation management tools
        -   supports LaTeX syntax for example sentences and tables

-   `papaja` package for APA-formatted Rmarkdown documents

-   challenge: collaboration

    -   not all collaborators know these tools
    -   track changes not currently possible


# Setting up a project

-   tomorrow: hands-on
-   required installations/recent versions of:
    -   R
        -   version `4.4.0`, "Puppy Cup"
        -   check current version with `R.version`
        -   download/update: <https://cran.r-project.org/bin/macosx/>
    -   RStudio
        -   version `2023.12.1.402`, "Ocean Storm"
        -   Help \> Check for updates
        -   new install: <https://posit.co/download/rstudio-desktop/>

# Learning objectives 🏁 {.unnumbered .unlisted .uncounted}

Today we learned...

-   how to implement a project-oriented workflow in R ✅
-   how to organise folders and file names ✅
-   how to set up our first R project ✅

# References {.unlisted .unnumbered visibility="uncounted"}

---
nocite: |
---

::: {#refs custom-style="Bibliography"}
:::





